---
title: "R Notebook"
output: html_notebook
---

# Introduction

This R Notebook continues the analysis of Minnesota's Congressional District 3 election data from 2012 to 2020. We will:

- Load the combined shapefile.
- Explore and summarize the data.
- Visualize voting trends and turnout.
- Analyze congressional election results.


# Libraries

We start by loading the necessary libraries for spatial data manipulation and visualization.

```{r, message=FALSE}
# Load necessary libraries
library(sf)
library(dplyr)
library(tmap)
library(ggplot2)
```

# Loading the Combined Shapefile
We load the combined shapefile for District 3 that was created in the previous steps.

```{r}
# Define the path to the combined shapefile
output_path <- "C:/Users/Norah Alotaibi/Downloads/shp_bdry_electionresults_2012_2020/district_3_combined_2012_2020.shp"

# Load the combined shapefile
district_3_sf <- st_read(output_path)
```

# Exploring the Data
We explore the structure and contents of the dataset to understand the available variables.

```{r}
# View structure of the data
str(district_3_sf)

# View the first few rows
head(district_3_sf)

# Summary of Total Voting
summary(district_3_sf$TOTVOTING)
```
_________________________________________________________________________
### Data Preparation

#Add Turnout Rate
We calculate the voter turnout rate for each precinct.

```{r}
# Calculate voter turnout rate
district_3_sf <- district_3_sf %>%
  mutate(Turnout_Rate = (TOTVOTING / REG7AM) * 100)
```

#Summarize Data
We summarize key metrics, including total votes, turnout rates, and party performance, for each year.
```{r}
# Summarize data by year
district3_summary <- district_3_sf %>%
  group_by(Year) %>%
  summarize(
    Total_Voters = sum(TOTVOTING, na.rm = TRUE),
    Total_Registered = sum(REG7AM, na.rm = TRUE),
    Average_Turnout_Rate = mean(Turnout_Rate, na.rm = TRUE),
    Republican_Votes = sum(USREPR, na.rm = TRUE),
    Democratic_Votes = sum(USREPDFL, na.rm = TRUE),
    WriteIn_Votes = sum(USREPWI, na.rm = TRUE)
  )

# View the summarized data
print(district3_summary)
```
_________________________________________________________________________

### Visualization

#Mapping Precinct-Level Voting
We create an interactive map to visualize total voting by precinct using the tmap package.
```{r}
# Set tmap to interactive view mode
tmap_mode("view")

# Interactive map showing total voting by precinct
tm_shape(district_3_sf) +
  tm_polygons("TOTVOTING", palette = "viridis", title = "Total Voting") +  
  tm_layout(title = "District 3 Precinct-Level Voting")  
```


#Total Voting Trends
We visualize the total number of votes cast over the years.
```{r}
# Plot total voting trends
ggplot(district3_summary, aes(x = Year, y = Total_Voters)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Total Voting Trends in District 3 (2012-2020)",
    x = "Year",
    y = "Total Votes"
  ) +
  theme_minimal()
```

#Party Performance Over Time

We compare the performance of Republican, Democratic, and Write-In candidates.

```{r}
# Plot party performance
ggplot(district3_summary, aes(x = Year)) +
  geom_line(aes(y = Republican_Votes, color = "Republican"), size = 1.2) +
  geom_line(aes(y = Democratic_Votes, color = "Democratic"), size = 1.2) +
  geom_line(aes(y = WriteIn_Votes, color = "Write-In"), size = 1.2) +
  geom_point(aes(y = Republican_Votes, color = "Republican"), size = 3) +
  geom_point(aes(y = Democratic_Votes, color = "Democratic"), size = 3) +
  geom_point(aes(y = WriteIn_Votes, color = "Write-In"), size = 3) +
  labs(
    title = "Congressional Election Results in District 3 (2012-2020)",
    x = "Year",
    y = "Votes",
    color = "Candidate"
  ) +
  theme_minimal()
```

#Turnout Rate Over Time
We visualize the average voter turnout rate for each year.
```{r}
# Plot average turnout rate
ggplot(district3_summary, aes(x = Year, y = Average_Turnout_Rate)) +
  geom_line(size = 1.2, color = "blue") +
  geom_point(size = 3, color = "blue") +
  labs(
    title = "Average Turnout Rate in District 3 (2012-2020)",
    x = "Year",
    y = "Average Turnout Rate (%)"
  ) +
  theme_minimal()
```

#Export Summarized Data
Finally, we export the summarized data for further analysis or sharing.
```{r}
write.csv(district3_summary, "district3_summary.csv", row.names = FALSE)
```


```{r}
install.packages("spdep")
```

#Identify Missing Values
Check for missing values in the Turnout_Rate column:

```{r}
sum(is.na(district_3_sf$Turnout_Rate))
```

#Remove Rows with Missing Values
If you decide that rows with missing values in Turnout_Rate are not critical for your analysis, you can remove them:
```{r}
district_3_sf <- district_3_sf %>%
    filter(!is.na(Turnout_Rate))
```


#Check for Non-Finite Values
Identify if there are any non-finite values in Turnout_Rate:
```{r}
sum(!is.finite(district_3_sf$Turnout_Rate))
```

#Filter Out Non-Finite Values
Remove any non-finite values from Turnout_Rate:
```{r}
district_3_sf <- district_3_sf %>%
    filter(is.finite(Turnout_Rate))
```


#Spatial Analysis:
Cluster Analysis: Identify clusters of high and low voter turnout using spatial clustering methods.
```{r}
library(spdep)
coords <- st_coordinates(st_centroid(district_3_sf))
nb <- knn2nb(knearneigh(coords, k = 5))
lw <- nb2listw(nb, style = "W")
district_3_sf$Moran_I <- localmoran(district_3_sf$Turnout_Rate, lw)[, "Ii"]

ggplot(district_3_sf, aes(fill = Moran_I)) +
    geom_sf() +
    scale_fill_viridis_c() +
    labs(title = "Spatial Clusters of Voter Turnout")
```

```{r}
summary(district_3_sf$Turnout_Rate)
```


```{r}
# Calculate percentage of votes for each party
district_3_sf <- district_3_sf %>%
  mutate(
    Percent_Democratic = (USREPDFL / TOTVOTING) * 100,
    Percent_Republican = (USREPR / TOTVOTING) * 100
  )

# Visualize percentage of votes
ggplot(data = district_3_sf) +
  geom_sf(aes(fill = Percent_Democratic), color = NA) +
  scale_fill_gradient(low = "white", high = "blue", name = "Democratic (%)") +
  labs(
    title = "Election Results by Percentage in MN Congressional District 03",
    subtitle = "Percentage of Votes Received by Democratic Candidates",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

# For Republican Percentage
ggplot(data = district_3_sf) +
  geom_sf(aes(fill = Percent_Republican), color = NA) +
  scale_fill_gradient(low = "white", high = "red", name = "Republican (%)") +
  labs(
    title = "Election Results by Percentage in MN Congressional District 03",
    subtitle = "Percentage of Votes Received by Republican Candidates",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

```








When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
