---
title: "R Notebook"
output: html_notebook
---

## Introduction

This R Notebook analyzes election results for Minnesota's Congressional District 3 using shapefiles from 2012 to 2020. The goal is to combine data, filter for District 3, and perform further analysis.


```{r}
library(sf)
library(dplyr)
```

#Loading Data
We define the path to the shapefiles and create a list of years for which we have data.


```{r}
# List of shapefiles by year
years <- c(2012, 2014, 2016, 2018, 2020)
base_path <- "C:/Users/Norah Alotaibi/Downloads/shp_bdry_electionresults_2012_2020"
```

#Extracting Column Names
To ensure consistency, we extract and find the common columns across all shapefiles.

```{r}

# Extract column names for each shapefile
column_names <- lapply(all_shapefiles, colnames)

# Find common columns
common_cols <- Reduce(intersect, column_names)

```

```{r}
# Keep only common columns in each shapefile
all_shapefiles <- lapply(all_shapefiles, function(gdf) gdf[, common_cols, drop = FALSE])

```

#Combining Shapefiles
We combine all the shapefiles into a single dataset.

```{r}
combined_sf <- do.call(rbind, all_shapefiles)

```

#Filtering for District 3
We filter the combined dataset to include only Congressional District 3.

```{r}
# Filter for District 3
district_3_sf <- combined_sf %>% filter(CONGDIST == 3)

# Save the combined dataset
output_path <- file.path(base_path, "district_3_combined_2012_2020.shp")
st_write(district_3_sf, output_path)

```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
